---
title: "Notes and Exercises"
author: "Shawn Garbett"
date: "7/24-26/2018"
output: html_document
---

```{r setup, include=FALSE}
library(msm)
library(tangram)
library(dplyr)
library(ggplot2)
library(dampack)
library(scales)
library(ellipse)
library(truncnorm)
library(reshape2)
library(plotrix)
library(lhs)
library(matrixStats)
knitr::opts_chunk$set(echo = TRUE)
source("Functions.R")
```

# Day 2

## Exercise: Probabilistic sensitivity analysis of the Sick-Sicker Markov model

This exercise continues based on the time-homogeneous deterministic Markov
model of the Sick-Sicker model from exercise “Exercise-
SickSickerMarkovModel.docx”. In this exercise, you will do a probabilistic sensitivity
analysis (PSA) with 1000 simulations (n.sim). The Table describes the distributions
for the variables you used in the previous exercise.

### Tasks

1. Use the R function of the Sick-Sicker Markov model in the file “Markov_Sick-
Sicker - SA – Function.R”.
2. Generate values for the uncertain parameters using the appropriate
distributions. *Hint: package truncnorm deals with truncated normal distributions.*
3. Conduct a probabilistic Cost-Effectiveness analysis of treatment vs no-
treatment
4. Create histograms of model inputs.
5. Create a cost-effectiveness plane to present discounted costs and QALYs.
6. Create a cost-effectiveness acceptability curve for the treatment comparison
assuming WTP thresholds of $0 to $200,000.
7. Create an expected value of perfect information (EVPI) plot. Expected Value
of Perfect Information (EVPI)

```{r}
rm(list=ls())
source('Functions.R')

d.r <- 0.03
n.t <- 30

hrf      <- function(p, h) 1-exp(log(1-p)*h)
discount <- 1 / (1 + d.r) ^ (0:n.t)

# Markov Model solver
markov <- function(n.t, m, i)
{
  y <- unlist(Reduce(function(x, y) x %*% m, 1:n.t, i, accumulate=TRUE))
  y <- matrix(y, ncol=dim(m)[1], byrow=TRUE)
  colnames(y) <- colnames(m)
  y
}

# Sick Sicker Markov Model
MM.SickSicker <- function(params)
{
  v.n    <- c("Healthy", "Sick", "Sicker", "Dead")
  with(as.list(params), {
    cohort.tr <- matrix(
      c(
        0,     p.HS1,     0,      p.HD,
        p.S1H,     0,     p.S1S2, hrf(p.HD,hr.S1D), 
        0,         0,     0,      hrf(p.HD,hr.S2D),
        0,         0,     0,      0
      ), 
      dimnames=list(v.n, v.n),
      nrow=length(v.n), byrow=TRUE
    )
    diag(cohort.tr)     <- 1 - rowSums(cohort.tr)
    markov(n.t, cohort.tr, c(1, 0, 0, 0))
  })
}
  
# Expectation of Sick Sicker Model
Expect.SickSicker <- function(params)
{
  solution <- MM.SickSicker(params)
  
  with(as.list(params), {
    # No Treat
    tc_no_trt   <- solution %*% c(c.H, c.S1, c.S2, c.D)
    tu_no_trt   <- solution %*% c(u.H, u.S1, u.S2, u.D)
    d_tc_no_trt <- t(tc_no_trt) %*% discount
    d_tu_no_trt <- t(tu_no_trt) %*% discount
    
    # Treat
    tc_trt   <- solution %*% c(c.H, c.S1+c.Trt, c.S2+c.Trt, c.D) 
    tu_trt   <- solution %*% c(u.H, u.Trt, u.S2, u.D) 
    d_tc_trt <- t(tc_trt) %*% discount
    d_tu_trt <-  t(tu_trt) %*% discount  
    
    c(d_cost_no_trt = d_tc_no_trt,
      d_qaly_no_trt = d_tu_no_trt,
      d_cost_trt    = d_tc_trt,
      d_qaly_trt    = d_tu_trt)
  })
}

### Function to generate PSA input dataset
gen_psa <- function(n = 1000, seed = 071818)
{
  set.seed(seed)
  ### Dataframe of input parameters
  data.frame(
    ### Probabilities
    p.HS1  = rbeta(n, 30,  170),         # Disease Onset Probability / year       (H  -> S1)
    p.S1H  = rbeta(n, 60,   60),         # Recovery Probability / year            (S1 -> H)
    p.S1S2 = rbeta(n, 84,  716),         # Disease Progression Probability / year (S1 -> S2)
    p.HD   = rbeta(n, 10, 1990),         # Healthy to Dead probability / year     (H  -> D)
    hr.S1D = rlnorm(n, log( 3), 0.01),   # Hazard ratio in S1 vs healthy 
    hr.S2D = rlnorm(n, log(10), 0.02),   # Hazard ratio in S2 vs healthy
    
    ### Annual Costs
    c.H    = rgamma(n, 100, 1/20),       # Healthy individuals 
    c.S1   = rgamma(n, 1600/9, 4/90),    # Sick individuals in S1
    c.S2   = rgamma(n, 255, 3/200),      # Sick individuals in S2
    c.Trt  = rgamma(n, 73.46939, 3/490), # Additional Annual cost for S1 and S2
    c.D    = 0,                          # Dead individuals

    # Utility Weights
    u.H    = rtruncnorm(n, mean=1,    sd=0.01, b=1),   # Healthy
    u.S1   = rtruncnorm(n, mean=0.75, sd=0.02, b=1),   # S1
    u.S2   = rtruncnorm(n, mean=0.5,  sd=0.03, b=1),   # S2
    u.D    = 0,                                        # Dead

    # Intervention Effect
    u.Trt  = rtruncnorm(n, mean=0.95, sd=0.02, b=1)    # S1 Utility for treatment in S1
  )
}

n.sim <- 1000
n.str <- 2 # Number of strategies

m.psa.input <- gen_psa(n.sim)
ggplot(melt(m.psa.input, variable.name = "Parameter"), aes(x = value)) +
  facet_wrap(~Parameter, scales = "free") +
  geom_histogram(aes(y = ..density..), bins=30) +
  theme_bw(base_size = 8)
```

```{r}
strategies <- c("No Treatment", "Treatment")
psa   <- t(apply(m.psa.input, 1, Expect.SickSicker))
costs <- psa[,c(1, 3)]
qalys <- psa[,c(2, 4)]
colnames(costs) <- colnames(qalys) <- strategies
ScatterCE(strategies = strategies, m.c = costs, m.e = qalys)
```

```{r}
wtp <-  seq(0, 200000, by = 10000)
ceaf(v.wtp = wtp, strategies = strategies, m.e = qalys, m.c = costs)$gg.ceaf
```

```{r}
plot(evpi(v.wtp = wtp, m.e = qalys, m.c = costs))
```

## Exercise: Calibration of the Sick-Sicker Model

As you may recall, the Sick-Sicker model simulates the progression of a cohort of
healthy 25-year-olds at risk of developing a hypothetical disease with two stages of
illness: a Sick state (S1) and a Sicker state (S2). Individuals in S1 have increased
mortality, increased healthcare costs, and decreased quality of life relative to those
in the Healthy state. Individuals in S2 have even greater increases in mortality,
costs, and quality of life reductions.

Previously, you were provided with a full set of input parameters for this model.
However, “in reality” some of those input parameters can’t feasibly be estimated
directly from individual-level data and need to be estimated through calibration.
This is because while we can identify those who are afflicted with the illness
through obvious symptoms, we can’t easily distinguish those in the S1 state from
the those in the S2 state. Thus, we can’t directly estimate state-specific mortality
hazard ratios, nor do we know the transition probability of progressing from S1 to
S2.

There are three calibration targets to be used in this example: (1) observed
survival of a cohort over time; (2) disease prevalence in a cohort over time; and
(3) among sick patients (S1+S2), the proportion who are in the S1 state at three
points in time. The first two targets came from an observational study of a typical
cohort of individuals. The third target was obtained by subjecting ill patients to an
invasive procedure which can determine disease stage, but can’t be used routinely
due to its invasiveness.

The calibration target values are provided in the data file
“SickSickerTargets.RData”, which contains a list with three data frames: Surv, Prev,
and PropSick.
The tasks below will guide you through the process of calibrating the Sick-Sicker
model to estimate values for the unknown input parameters.

* Adapt the function of the Sick-Sicker Markov model R code in “Markov_Sick-
Sicker - SA - Function.R” so that the model returns model outputs that
correspond to the calibration targets. Make this model into a function that
can be called in the calibration process.
* Load the calibration targets and plot them over time.
* Implement a random search of the unknown parameters using Latin
hypercube sampling. Plot a histogram of the sampled parameter values.
* Run the Sick-Sicker Markov model for each of the sample parameter sets and
store the relevant model outputs.
* Calculate the fit of each model output to the calibration targets using a
normal likelihood goodness-of-fit and identify the best-fitting parameter set.
* Plot the model outputs corresponding to the best-fitting parameter set
against each calibration target. Visually, is the model well-calibrated to the
targets?

```{r}
load("SickSickerTargets.RData")
n.t <- 30
par(mfrow=c(1,3))
plotrix::plotCI(x = 1:n.t, y = l.targets$Surv$value, 
                ui = l.targets$Surv$ub,
                li = l.targets$Surv$lb,
                ylim = c(0, 1), 
                xlab = "Time", ylab = "Survival")

plotrix::plotCI(x = 1:n.t, y = l.targets$Prev$value, 
                ui = l.targets$Prev$ub,
                li = l.targets$Prev$lb,
                ylim = c(0, 1), 
                xlab = "Time", ylab = "Prevalance")

plotrix::plotCI(x = l.targets$PropSick$Time, y = l.targets$PropSick$value, 
                ui = l.targets$PropSick$ub,
                li = l.targets$PropSick$lb,
                ylim = c(0, 1), 
                xlab = "Time", ylab = "Proportion Sick")
```

```{r}

ll.SickSicker <- function(params, summary=FALSE)
{
  solution   <- MM.SickSicker(params)
  
  survival   <- 1-solution[2:(n.t+1),4]
  prevalance <- rowSums(solution[2:(n.t+1),c(2,3)]) / survival
  sick       <- solution[c(11, 21, 31), 2] / prevalance[c(10, 20, 30)]
  
  ll <-
    sum(dnorm(x    = l.targets$Surv$value, 
              mean = survival, 
              sd   = l.targets$Surv$se, 
              log  = TRUE)) +
    sum(dnorm(x    = l.targets$Prev$value, 
              mean = prevalance, 
              sd   = l.targets$Prev$se, 
              log  = TRUE)) +
    sum(dnorm(x    = l.targets$PropSick$value, 
              mean = sick, 
              sd   = l.targets$PropSick$se, 
              log  = TRUE))
  
  if(summary)
  {
    attr(ll, "survival")   <- survival
    attr(ll, "prevalance") <- prevalance
    attr(ll, "sick")       <- sick
  }
  
  ll
}

gen_lhs <- function(n = 10000, seed = 071918)
{
  set.seed(seed)
  lhs.samp <- randomLHS(n = n, k = 3)
  
  data.frame(
    ### Probabilities
    p.HS1  =     0.15,            # Disease Onset Probability / year       (H  -> S1)
    p.S1H  =     0.5,             # Recovery Probability / year            (S1 -> H)
    p.S1S2 = qunif(lhs.samp[,1], 0.01, 0.5), # Disease Progression Probability / year (S1 -> S2)
    p.HD   =     0.005,           # Healthy to Dead probability / year     (H  -> D)
    hr.S1D = qunif(lhs.samp[,2], 1, 4.5),   # Hazard ratio in S1 vs healthy 
    hr.S2D = qunif(lhs.samp[,3], 5, 15)     # Hazard ratio in S2 vs healthy
  )
}

lhs.grid <- gen_lhs()
cal   <- apply(lhs.grid, 1, ll.SickSicker)
hist(cal, freq=TRUE, main="Calibration Likelihoods")
```

```{r}
mll.lhs <- lhs.grid[which.max(cal),]
mll.lhs
```

```{r}
solution <- ll.SickSicker(mll.lhs, summary=TRUE)

par(mfrow=c(1,3))
plotrix::plotCI(x = 1:n.t, y = l.targets$Surv$value, 
                ui = l.targets$Surv$ub,
                li = l.targets$Surv$lb,
                ylim = c(0.5, 1), 
                xlab = "Time", ylab = "Pr Survive")
lines(x=1:n.t, attr(solution, "survival"), col='red', lty=2, lwd=2)

plotrix::plotCI(x = 1:n.t, y = l.targets$Prev$value, 
                ui = l.targets$Prev$ub,
                li = l.targets$Prev$lb,
                ylim = c(0, 1), 
                xlab = "Time", ylab = "Prevalance")
lines(x=1:n.t, attr(solution, "prevalance"), col='red', lty=2, lwd=2)

plotrix::plotCI(x = l.targets$PropSick$Time, y = l.targets$PropSick$value, 
                ui = l.targets$PropSick$ub,
                li = l.targets$PropSick$lb,
                ylim = c(0, 1), 
                xlab = "Time", ylab = "Proportion Sick")
lines(x=c(10, 20, 30), attr(solution, "sick"), col='red', lty=2, lwd=2)
```

```{r}
slice <- function(param1, a1, b1,
                  param2, a2, b2,
                  levels, 
                  optim,
                  n=50)
{
  x <- seq(a1, b1, by=(b1-a1)/(n-1))
  y <- seq(a2, b2, by=(b2-a2)/(n-1))
  z <- matrix(data=NA, nrow=50, ncol=50)
  
  for(i in 1:50)
  {
    for(j in 1:50)
    {
      ref <- optim
      ref[param1] <- x[i]
      ref[param2] <- y[j]
      z[i,j]      <- ll.SickSicker(ref)
    }
  }

  contour(x, y, z, levels=lines, xlab=param1, ylab=param2)
  points(optim[param1], optim[param2], pch=2, col="red", cex=2)
  legend("topright", "LHS MLE", col="red", pch=2)
}

lines <- round(quantile(cal, seq(0.1, 0.9, by=0.1)))

par(mfrow=c(1,3))
slice("p.S1S2", 0.01,  0.5,
      "hr.S1D", 1,     4.5,
       lines, mll.lhs)

slice("p.S1S2", 0.01,  0.5,
      "hr.S2D", 5,    15,
       lines, mll.lhs)

slice("hr.S1D", 1,     4.5,
      "hr.S2D", 5,    15,
       lines, mll.lhs)

```

Thus concluding Day 2