---
title: "Notes and Exercises"
author: "Shawn Garbett"
date: "7/24-26/2018"
output: html_document
---

```{r setup, include=FALSE}
library(msm)
library(tangram)
library(dplyr)
library(ggplot2)
library(dampack)
library(scales)
library(ellipse)
library(truncnorm)
library(reshape2)
library(plotrix)
library(lhs)
library(matrixStats)
knitr::opts_chunk$set(echo = TRUE)
source("Functions.R")
```

# Day 3

## Exercise: Variations on the Sick-Sicker Markov Model

Previously, you built a Markov of the Sick-Sicker model where transition
probabilities were assumed to be constant over time. In this exercise, you will
expand on that model to incorporate age-dependence (time-varying probabilities)
and history-dependence.

### Part 1: Age-dependence

Healthy individuals are no longer assumed to have a fixed mortality rate. Their
mortality rate depends on their age. The Human Mortality Database (HMD)
(www.mortality.org) provides these age specific mortality rates (Mx1x1). You can
load this data in R using different packages, for example the HMDHFplus,
demography or data.table package. Registration (free) on the HMD website is
required in order to use the data. For this exercise, we provide you with this
mortality rate for the US in the file “HMD_USA_Mx_2015.csv”.
Remember: individuals in S1 and S2 still have an increased mortality relative to
healthy individuals, as described in the original exercise. The same hazard ratios
are used to calculate the probabilities of dying from S1 and S2 in this exercise.

* Use the R function “Markov_Sick-Sicker.R” and create a copy to then adapt
the model to account for age-dependency.
* Import the mortality rate data the “HMD_USA_Mx_2015.csv” file and store
this data in a variable called v.r.HD.
* Create probability to die when healthy as a function of age `p.HD <- 1 - exp(- v.r.HD*a)`
* Incorporate the age dependent mortality rate in the model.
* Create a survival curve of the cohort under no treatment. *Extra challenge: plot the survival curves with and without age-dependent probabilities of death in one graph so you can compare.*

```{r}
hrf      <- function(p, h) 1-exp(log(1-p)*h)

# Sick Sicker Markov Model
MM.SickSicker <- function(params)
{
  v.n    <- c("Healthy", "Sick", "Sicker", "Dead")
  with(as.list(params), {
    cohort.tr <- matrix(
      c(
        0,     p.HS1,     0,      p.HD,
        p.S1H,     0,     p.S1S2, hrf(p.HD,hr.S1D), 
        0,         0,     0,      hrf(p.HD,hr.S2D),
        0,         0,     0,      0
      ), 
      dimnames=list(v.n, v.n),
      nrow=length(v.n), byrow=TRUE
    )
    diag(cohort.tr)     <- 1 - rowSums(cohort.tr)
    markov(n.t, cohort.tr, c(1, 0, 0, 0))
  })
}

# Markov Model solver with possible time dependent modification function
# Defaults to identity of matrix function
markov <- function(n.t, m, i, FUN=function(mm, aa, p) mm, params=NULL)
{
  y <- unlist(Reduce(function(x, y) x %*% FUN(m, y, params), 1:n.t, i, accumulate=TRUE))
  y <- matrix(y, ncol=dim(m)[1], byrow=TRUE)
  colnames(y) <- colnames(m)
  y
}

# Load Mortality Data
lt.usa.2005 <- read.csv("HMD_USA_Mx_2015.csv")

# Target population is age 25
td.p.HD <- 1-exp(-lt.usa.2005[lt.usa.2005$Age >= 25,'Total'])
#td.p.HD <- lt.usa.2005[lt.usa.2005$Age >= 25,'Total']

# This function takes a transition probability matrix and current time
# Changes the mortality probability to be time based from 
mortality_ad <- function(m, a, params)
{ 
  m["Healthy", "Dead"] <- td.p.HD[a]
  m["Sick",    "Dead"] <- hrf(td.p.HD[a],params$hr.S1D)
  m["Sicker",  "Dead"] <- hrf(td.p.HD[a],params$hr.S2D)
  diag(m) <- 0
  diag(m) <- 1 - rowSums(m)
  m
}

# Sick Sicker Markov Model with Age Dependence
MMAG.SickSicker <- function(params)
{
  v.n    <- c("Healthy", "Sick", "Sicker", "Dead")
  with(as.list(params), {
    cohort.tr <- matrix(
      c(
        0,     p.HS1,     0,      p.HD,
        p.S1H,     0,     p.S1S2, hrf(p.HD,hr.S1D), 
        0,         0,     0,      hrf(p.HD,hr.S2D),
        0,         0,     0,      0
      ), 
      dimnames=list(v.n, v.n),
      nrow=length(v.n), byrow=TRUE
    )
    diag(cohort.tr)     <- 1 - rowSums(cohort.tr)
    markov(n.t, cohort.tr, c(1, 0, 0, 0), mortality_ad, params)
  })
}

params <- list(
  n.t    =    30,
  
  p.HS1  =     0.15,   # Disease Onset Probability / year       (H  -> S1)
  p.S1H  =     0.5,    # Recovery Probability / year            (S1 -> H)
  p.S1S2 =     0.105,  # Disease Progression Probability / year (S1 -> S2)
  p.HD   =     0.005,  # Healthy to Dead probability / year     (H  -> D)
  hr.S1D =     3,      # Hazard ratio in S1 vs healthy 
  hr.S2D =    10,      # Hazard ratio in S2 vs healthy,
  
  # Annual Costs
  c.H    =  2000,      # Healthy individuals 
  c.S1   =  4000,      # Sick individuals in S1
  c.S2   = 15000,      # Sick individuals in S2
  c.D    =     0,      # Dead individuals
  c.Trt  = 12000,      # Additional Annual cost for S1 and S2,
  
  # Utility Weights
  u.H    =     1.00,   # Healthy
  u.S1   =     0.75,   # S1
  u.S2   =     0.50,   # S2
  u.D    =     0.00,   # Dead,
  
  # Intervention Effect
  u.Trt  =     0.95    # S1 Utility for treatment in S1
)

noage <- MM.SickSicker(params)
w_age <- MMAG.SickSicker(params)

noage_surv <- 1-noage[,"Dead"]
w_age_surv <- 1-w_age[,"Dead"]

plot(0:params$n.t+25, noage_surv, ylim=c(0,1),
     xlab="Age", ylab="Survival Probability",
     main="Overall Survival", typ='l')
grid(nx = params$n.t, ny = 10, col = "lightgray", lty = "dotted", lwd = par("lwd"), equilogs = TRUE)
lines(0:params$n.t+25, w_age_surv, lty=2, col='red')
legend("bottomleft", c("Time Homogeneous", "Age-Dependent"), col=c("black", "red"),
       lty=1:2)
```

### Part 2

It has been recently discovered that the risk of progression from Sick to Sicker
increases the longer a person has been sick. This increase follows a Weibull growth
curve. We will now expand the age-dependent model to include this history dependence by
adding tunnel states for S1.

* Starting from the age-dependent Markov model in Part 1, expand the 3D
transition probability array to account for tunnels
* Create the parameter p.S1S2 as a Weibull function as follows

```
p.S1S2 <- l*g*(1:tunnel.size)^{g-1}
l <- 0.08 # scale
g <- 1.1 # shape
```

* Fill in the 3D transition probability array accounting for the tunnel states for
S1
* Plot the survival curve for the cohort under no treatment. *Extra challenge: plot the survival curves for all three Markov model
versions (time-homogenous, age-dependent, and history-dependent)
in one graph so you can compare.*


```{r}

# Dynamic Markov Solver. Needs function to return each iteration
dynamic_markov <- function(n.t, i, FUN, params=NULL)
{
  y <- Reduce(function(x, y) x %*% FUN(y, params), 1:n.t, i, accumulate=TRUE)
  y <- matrix(unlist(y), ncol=length(y[[1]]), byrow=TRUE)
  colnames(y) <- colnames(FUN(1, params))
  y
}

tp.SickSicker <- function(a, params)
{
  with(as.list(params), {
    v.n    <- c("Healthy", paste0("S", 1:n.t), "Sicker", "Dead")
    trp <- matrix(ncol=n.t+3, byrow=TRUE,
                  dimnames=list(v.n, v.n),
                  c(
                    0,         rep(0, n.t),     0,      td.p.HD[a],             # Healthy -> 
                    rep(0, n.t*(n.t+3)),                                  # Sx -> 
                    0,         rep(0, n.t),     0,      hrf(td.p.HD[a],hr.S2D), # Sicker -> 
                    0,         rep(0, n.t),     0,      1                 # Dead (absorbing)
                  )
    )
    trp[1, 2] <- p.HS1
    p.SxD  <- hrf(td.p.HD[a],hr.S1D)
    for(i in (1:n.t) + 1)
    {
      S1S2 <- 0.08 * 1.1 * (i-1)^0.1
      trp[i, 1]        <- p.S1H # Sx -> H
      trp[i, "Sicker"] <- S1S2
      trp[i, "Dead"]   <- p.SxD
      if(i < n.t+1) trp[i, i+1]      <- 1-S1S2 - p.SxD - p.S1H
    }
    diag(trp) <- 0
    diag(trp)     <- 1 - rowSums(trp)
    
    trp
  })
}

# Sick Sicker Markov Model with Age Dependence and Tunnel State
MMTS.SickSicker <- function(params) 
  dynamic_markov(params$n.t, c(1, rep(0, params$n.t+2)), tp.SickSicker, params)


tunn_surv <- 1-MMTS.SickSicker(params)[,"Dead"]
first3    <- rowSums(MMTS.SickSicker(params)[,1:3])

plot(0:params$n.t+25, noage_surv, ylim=c(0,1),
     xlab="Age", ylab="Survival Probability",
     main="Overall Survival", typ='l')
grid(nx = params$n.t, ny = 10, col = "lightgray", lty = "dotted", lwd = par("lwd"), equilogs = TRUE)
lines(0:params$n.t+25, w_age_surv, lty=2, col='red')
lines(0:params$n.t+25, tunn_surv, lty=3, col='blue', lwd=2)
lines(0:params$n.t+25, first3, lty=4, col='darkgreen', lwd=1)
legend("bottomleft", c("Time Homogeneous", "Age-Dependent", "Tunnels", "H+S1+S2"), col=c("black", "red", "blue", "darkgreen"),
       lty=1:4, lwd=c(1,1,2,1))
```
